---
title: "Mileage Analysis of Auto & Manual Cars (mtcars)"
author: "AK"
date: "April 5, 2016"
output: html_document
runtime: shiny
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Application: Overview
This R Markdown document is made interactive using Shiny. It enables the user to perform analysis of
mileage (mpg) for automatic and manual transmission cars. Employing regression modeling, it
enables user to explore the relationship between fuel efficiency and a set of variables in mtcars..
This is shinification of the project I did for the project for the Regression Modeling Course.
### Application: Inputs and Outputs
The user can choose one of the four regression models using the radio buttons on the left. A summary of relationships between all the variables in the mtcars dataset has been presented in the section below these options. The application displays the following information on the right for the chosen model:

- Summary Statistics for the Chosen Model (Tab 1)
- Plot for the Chosen Model (Tab 2)

It also presents 

- Pairwise Plot for the Significant Variables (Tab 3)
- MPG vs Transmission (Tab 4)

The user can compare the regression model output to conclude the desired model to use.

Guidelines for performig the Analysis:
- Review the pair-wise plot to understand the variable relationships
- Review the MPG vs Transmission plot
- Run each regression model
- Carefully review the R-Squared values for each model and compare
- Choose the model with the highest R-Sqaured value

I have embedded the entire AKMileageAnalysisShinyApp application within this R Markdown document. Please scroll down/click on the link to launch the application and play with it. 
```{r embeddedapp, echo = FALSE, results='asis'}

# Here is a direct link to the AKMileageAnalyisisShinyApp on shinyapps.io
cat('<br>')
cat('<br>')
cat('<a href="https://akoratkar.shinyapps.io/AKMileageAnalysisShinyApp/">Mileage Analysis Shiny App on shinyapps.io</a>')

# Ignore below since the app embedding works on the local machine only but does not when uploaded to shinyapps. Leaving the code to future reference.

# shinyAppDir(
# "AKMileageAnalysisShinyApp",
# options=list(
#   width="100%", height="800"
#  )
# )

```

Don't peek! :)

The following sections presents the author's analysis and conclusion. The application user should do her own analysis and conclude based on the output presented.

### Author's Analysis

- Model: lm(mpg~am)
Mean MPG for Auto Transmission is 17.1473684 and Mean MPG for Manual Transmission is 24.3923077
The value of R-Squared indicates that only 35.9798943 % regression variance is explained by the lm(mpg~am) model. Hence, one should consider other variables that are included in the mtcars data set for more precise analysis.

- Models: lm(mpg~wt+am), lm(mpg~disp+am) and lm mpg~cyl+am)
The above models indicate that when wt and disp variables are considered separately, appendng the am does not improve the model significantly (smaller F changes), but there is some change (at 0.05) when we consider cyl with am.

- Model: lm(mpg~wt+cyl+am)
R-Squared indicates that a high 83.0338317 % regression variance is explained by the model. We conclude wt and cyl are significant contributors to mpg and lm(mpg~wt+cyl+am, data=mtcars) is the most optimal model.

### Author's Conclusion
Per the basic linear regression model that only considers the relationship between mpg and transmission type, the mean mpg for cars with automatic transmission is 17.147 and that for manual transmission is 24.392, with a mean difference of 17.147. Thus, excluding the effect of other variables, the manual transmission cars have better fuel efficiency (mileage) than the automatic tranmission cars.

Per the most optimal multi-variable linear regression model that considers the relationshiop between mpg and a set of other variables like weight, number of cylinders along with transmission type, the mpg difference between manual and automatic reduces to 0.1765 mpg at the mean weight and cyl, still. Thus, when other variables considered, manual transmission cars have marginally better fuel efficiency than the automatic cars.






