---
title       : Mileage Analysis of Auto & Manual Cars 
subtitle    : Shiny Application for Regression Modeling for Mileage Analysis with mtcars
author      : AK
job         : Life Long Learner
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : standalone # {standalone, selfcontained, codraft}
knit        : slidify::knit2slides

---

## Objective & Features of the Shiny Application

- The Shiny Appication enables the user to analyze the mtcars dataset and compare the fuel efficiency of the automatic and manual transmission cars.

- Left Pane: The user can choose from multiple linear regression models (radio buttons)

- Right Pane: The App displays the following to help user select the optimal model

  - Tab 1: Model Summary (chosen model)
  - Tab 2: Model Plot (chosen model)
  - Tab 3: Pair Wise Plot (for data set)
  - Tab 4: MPG vs Transmission Plot
- Bottom Pane: The App displays variance table ifor analyzing multivariable nested models.
  
- <a href="https://akoratkar.shinyapps.io/AKMileageAnalysisShinyApp/">Link to Mileage Analysis Shiny App on shinyapps.io</a>
- <a href="https://akoratkar.shinyapps.io/AKInteractiveMileageAnalysisShinyDocument/">Link to Mileage Analysis Shiny App R Markdown Documentation on shinyapps.io</a>

---


## Pair-wise Plot for mtcars
- This plot gives the user a high level view of the relationship among significant variables
```{r pairwiseplot, echo=FALSE, results="markup"}
  mtcars_wca <- mtcars[, c(1, 2, 6, 9)]
  pairs(mtcars_wca, panel=panel.smooth,col=11 + mtcars_wca$w, main="Plot 2: Pairwise plot for mpg, cyl, wt, am")

```

---

## Guidelines for analyzing the regression models
- Review the pair-wise plot and mpg vs am plot to understand the variable relationships
- Run each regression model. Carefully review the R-Squared values. E.g.
```{r samplersqr, echo=TRUE, results="markup"}
  summary(lm(mpg~wt+cyl+am, data=mtcars))$r.squared

```
- Review the variable table for the nested model. E.g.
- Choose the model with the highest R-Squared value
```{r sampleanova, echo=FALSE, results="markup"}
  head(summary(anova(lm(mpg~wt, data=mtcars), lm(mpg~wt+cyl, data=mtcars), lm(mpg~wt+cyl+disp, data=mtcars), lm(mpg~wt+cyl+disp+am, data=mtcars))), 2)

```

---

## Sample Model Output: lm(mpg~wt+cyl+am)
```{r samplemodel, echo=FALSE, results="markup"}
  
  summary
  
```



